Contracts for Contracts

### 2023-06-14 paper/contract_solidity.tex
misc

We begin with specifying precondition and postcondition for functions
involving first-order values. Consider the following \code{toEither} example
that converts \note{Wei?} to Either:
\begin{lstlisting}[language=Solidity]
# { toEither(x) returns (y)
    requires x % 1e18 < 1e17
    ensures  y < type(uint).max / 1e18 }
function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

@@

\paragraph{Syntactic Sugars}
We can liberally omit any part of the argument or return value specification:
\begin{lstlisting}
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18 }
function toWei(uint x) returns (uint) { ... }
\end{lstlisting}
which is equivalent to core form where the omitted part is simply the boolean expression
\code{true}:
\begin{lstlisting}[language=Solidity]
# { toWei(x) returns (_)
    ensures x < type(uint).max / 1e18
    requires true }
\end{lstlisting}

@@

\paragraph{Any Expression is Allowed}
When specifying the pre-/postconditions of functions, the programmer can use
any valid Solidity expression to from the condition, including but not limited
to function calls, memory operations, and built-in special variables carrying
important transaction data (e.g. \code{msg.value}) or metadata (e.g.
\code{block.timestamp}) that are only available at runtime.
The programmers are free to check desired conditions against with these
transactional information.
For example, the following snippet examines \code{msg.value} as the
precondition of a payable function:
\begin{lstlisting}
# { byTickets(n) ensures msg.value >= 1e15 * n }
function buyTickets(int n) payable { ... }
\end{lstlisting}
With this kind of flexibility and expressiveness, programmers are able to
express invariants beyond the capability of static verification.
%\zz{How about other special variables, such as `block.chainid` and `tx.origin`?. Perhaps we could redesign a universal syntax for these variables in Section 2.5, where we discuss where to add the message sender}
%\url{https://docs.soliditylang.org/en/develop/units-and-global-variables.html#block-and-transaction-properties}

@@

\paragraph{Addresses}
Resembling pointers in C/C++, addresses in Solidity are numerical values
that can be compared or \note{computed?}.
In this sense, specifications of addresses are no different
from other integer values.
For instance, the following specification requires that
the address argument value cannot be 0:
\begin{lstlisting}
# { f(addr) requires addr == 0 }
function f(address payable addr) { ... }
\end{lstlisting}

However, addresses in Solidity have rich higher-order behaviors: they may
represent a deployed external Ethereum contract and can be invoked in
customized ways.
For example, the following snippet shows a function \code{f} taking an address
\code{addr} that is called at some point.
The programmer intends to the specify the precondition and postcondition when
calling \code{addr}. \note{GW: explain abi?}
Nevertheless, such checks against \code{addr} cannot be performed when \code{f}
is called, even though \code{addr} is a proper argument of \code{f}.
Because \code{addr} in this use case is effectively a function, and
it is undecidable to statically examine its properties in general.
Additionally, we may not have access to the source code of the contract
represented by \code{addr}.
\begin{lstlisting}
# f(addr) where
  { addr.call(msg, amount) returns (flag, data)
    requires amount > 100
    ensures flag == true }
function f(address payable addr) {
  ... // omitting statements before the call
  (bool success, bytes memory data) =
    addr.call(abi.encodeWithSignature(
        "foo(string, uint256)", "call foo", 128
      )
    );
  ... // omitting statements after the call
}
\end{lstlisting}
\lang addresses this issue by adopting lazy checks
\cite{DBLP:conf/icfp/FindlerF02} that only take place
when the address is actually invoked \note{GW: fwd ref to impl}.
The above snippet specifies the condition of \code{addr}
in the \code{where}-clause, which requires
the \code{amount} argument must be greater than 100 and
the address call must succeed.

@@

\paragraph{Multiple Targets}
Multiple callee targets can inhabit in a single address value.
Other than the \code{call} method, there are other forms of low-level call,
e.g., \code{delegatecall}, \code{send}, etc.
% https://docs.soliditylang.org/en/v0.8.20/units-and-global-variables.html#address-related
Programmers can specify conditions for different callee targets in the
\code{where}-clause.
\begin{lstlisting}
# f(addr) where
  { addr.send(x) returns (succ) requires ... }
  { addr.call(msg, x) returns (succ, data) requires ... }
\end{lstlisting}

>>>>

We begin with specifying precondition and postcondition for functions
involving first-order values. Consider the following \code{toEither} example
that converts \note{Wei?} to Either:
\begin{lstlisting}[language=Solidity]
# toEither(x) returns (y)
# requires x % 1e18 < 1e17
# ensures  y < type(uint).max / 1e18
function toEther(uint x) returns (uint y) { ... }
\end{lstlisting}

@@

\paragraph{Syntactic Sugars}
We can liberally omit any part of the argument or return value specification:
\begin{lstlisting}
# toWei(x) returns (_)
# ensures x < type(uint).max / 1e18
function toWei(uint x) returns (uint) { ... }
\end{lstlisting}
which is equivalent to core form where the omitted part is simply the boolean expression
\code{true}:
\begin{lstlisting}[language=Solidity]
# toWei(x) returns (_)
# ensures x < type(uint).max / 1e18
# requires true
\end{lstlisting}

@@

\paragraph{Any Expression is Allowed}
When specifying the pre-/postconditions of functions, the programmer can use
any valid Solidity expression to from the condition, including but not limited
to function calls, memory operations, and built-in special variables carrying
important transaction data (e.g. \code{msg.value}) or metadata (e.g.
\code{block.timestamp}) that are only available at runtime.
The programmers are free to check desired conditions against with these
transactional information.
For example, the following snippet examines \code{msg.value} as the
precondition of a payable function:
\begin{lstlisting}
# byTickets(n) ensures msg.value >= 1e15 * n
function buyTickets(int n) payable { ... }
\end{lstlisting}
With this kind of flexibility and expressiveness, programmers are able to
express invariants beyond the capability of static verification.
%\zz{How about other special variables, such as `block.chainid` and `tx.origin`?. Perhaps we could redesign a universal syntax for these variables in Section 2.5, where we discuss where to add the message sender}
%\url{https://docs.soliditylang.org/en/develop/units-and-global-variables.html#block-and-transaction-properties}

@@

\paragraph{Addresses}
Resembling pointers in C/C++, addresses in Solidity are numerical values
that can be compared or \note{computed?}.
In this sense, specifications of addresses are no different
from other integer values.
For instance, the following specification requires that
the address argument value cannot be 0:
\begin{lstlisting}
# f(addr) requires addr == 0
function f(address payable addr) { ... }
\end{lstlisting}

However, addresses in Solidity have rich higher-order behaviors: they may
represent a deployed external Ethereum contract and can be invoked in
customized ways.
For example, the following snippet shows a function \code{f} taking an address
\code{addr} that is called at some point.
The programmer intends to the specify the precondition and postcondition when
calling \code{addr}. \note{GW: explain abi?}
Nevertheless, such checks against \code{addr} cannot be performed when \code{f}
is called, even though \code{addr} is a proper argument of \code{f}.
Because \code{addr} in this use case is effectively a function, and
it is undecidable to statically examine its properties in general.
Additionally, we may not have access to the source code of the contract
represented by \code{addr}.
\note{GW: could use a simpler example, and discuss low-level call in the next paragraph section}
\begin{lstlisting}
# f(addr) where
# { addr.call(msg, amount) returns (flag, data)
#   requires amount > 100
#   ensures flag == true }
function f(address payable addr) {
  ... // omitting statements before the call
  (bool success, bytes memory data) =
    addr.call(abi.encodeWithSignature(
        "foo(string, uint256)", "call foo", 128
      )
    );
  ... // omitting statements after the call
}
\end{lstlisting}
\lang addresses this issue by adopting lazy checks
\cite{DBLP:conf/icfp/FindlerF02} that only take place
when the address is actually invoked \note{GW: fwd ref to impl}.
The above snippet specifies the condition of \code{addr}
in the \code{where}-clause, which requires
the \code{amount} argument must be greater than 100 and
the address call must succeed.

@@

\paragraph{Multiple Targets}
Multiple callee targets can inhabit in a single address value.
Other than the \code{call} method, there are other forms of low-level call,
e.g., \code{delegatecall}, \code{send}, etc.
% https://docs.soliditylang.org/en/v0.8.20/units-and-global-variables.html#address-related
Programmers can specify conditions for different callee targets in the
\code{where}-clause.
\begin{lstlisting}
# f(addr) where
# { addr.send(x) returns (succ) requires ... }
# { addr.call(msg, x) returns (succ, data) requires ... }
\end{lstlisting}
